\documentclass[aspectratio=169]{beamer}
\usetheme{Madrid}
\usecolortheme{default}
\usefonttheme{professionalfonts}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}

\setbeamertemplate{navigation symbols}{}

\title[NLP in Python]{\textbf{Natural Language Processing in Python}}
\subtitle{HSE University --- Master’s in Data Science }
\author[Shalileh]{Assoc.\ Prof.\ Soroosh Shalileh}
\institute[HSE \& Sber]{
\inst{2}Higher School of Economics (HSE University)
\and 
\inst{1} Executive director of SberIndex, SberBank
}
\date{Spring Semester (Modules 3--4)}

\begin{document}

%------------------------------------------------
\begin{frame}
  \titlepage
\end{frame}
%------------------------------------------------

\begin{frame}{1.~Course Motivation and Context}
\begin{itemize}
  \item \textbf{Why NLP now?} Explosive growth of language data and large language models (LLMs) transforming research and industry.
  \item \textbf{Goal:} Equip students with the theoretical foundations and practical skills to design, train, and evaluate NLP systems.
  \item \textbf{Fit within program:} Advanced elective for Data Science \& AI focusing on text analytics, deep learning, and generative models.
  \item \textbf{Main reference:} Jurafsky \& Martin, \emph{Speech and Language Processing (3rd~ed.\ draft, 2025)}.
\end{itemize}
\end{frame}
%------------------------------------------------

\begin{frame}{2.~Course Structure}
\begin{columns}[T]
\column{0.48\textwidth}
\textbf{Module 3: Foundations}
\begin{itemize}
  \item Tokenization, normalization
  \item N-gram language models
  \item Logistic regression \& text classification
  \item Word embeddings \& neural networks
  \item Mini-Project: Classical/embedding pipeline
\end{itemize}

\column{0.48\textwidth}
\textbf{Module 4: Advanced Topics }
\begin{itemize}
  \item Transformers and BERT
  \item Retrieval-Augmented Generation (RAG)
  \item Instruction tuning \& alignment
  \item Machine translation, interpretability (if time permist)
  \item Final Project: research or applied NLP system
\end{itemize}
\end{columns}
\end{frame}
%------------------------------------------------

\begin{frame}{3.~Learning Outcomes}
By the end of the course, students will be able to:
\begin{enumerate}[leftmargin=*]
  \item Implement and compare classical and neural NLP models in Python.
  \item Fine-tune and evaluate transformer architectures using \texttt{PyTorch} and \texttt{HuggingFace}.
  \item Build retrieval-augmented and instruction-tuned systems.
  \item Critically analyze performance, fairness, and interpretability.
  \item Communicate findings through reproducible experiments and concise reports.
\end{enumerate}
\end{frame}
%------------------------------------------------

\begin{frame}{4.~Evaluation and Projects}
\begin{itemize}
  \item \textbf{Homeworks (7)} --- 35\%: implementation + analysis.
  \item \textbf{Mini-Project} --- 20\%: foundational NLP pipeline.
  \item \textbf{Final Project} --- 30\%: research or applied system.
  \item \textbf{Participation \& quizzes} --- 15\%.
\end{itemize}

\medskip
\textbf{Project Examples:}
\begin{itemize}
  \item Domain-specific BERT fine-tuning (legal or finance)
  \item Cross-lingual sentiment analysis
  \item Retrieval-augmented QA
  \item Instruction-tuning experiment on compact models
\end{itemize}
\end{frame}
%------------------------------------------------

\begin{frame}{5.~Takeaways and Contact}
\begin{itemize}
  \item Hands-on, research-oriented exploration of NLP fundamentals to advanced transformer methods.
  \item Encourages creativity, reproducibility, and critical analysis.
  \item Builds bridges between data science practice and modern AI research.
\end{itemize}

\bigskip
\textbf{Instructor:} Assoc.\ Prof.\ Soroosh Shalileh\\
\textbf{Email:} \texttt{sshalileh@hse.ru}; \\ 
\textbf{\ \ \ \ \ \ \ \ } sr.shalileh@gmail.com \\
\textbf{Office Hours:} By appointment\\[0.5em]
\textbf{Course page:} \url{https://web.stanford.edu/~jurafsky/slp3/} (textbook reference)

\textbf{Communication} {\color{red}fill-in the form using} \href{https://docs.google.com/forms/d/e/1FAIpQLScSjsQMZNR5WmPqVBbzrdmQPuR3xlC9FB6X2P1VQGl_u9ROWw/viewform?usp=publish-editor}{\color{blue}{this link}}
below to receive course materials.


\end{frame}
%------------------------------------------------
\end{document}
