\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{parskip}
\usepackage{amsmath, amssymb} 
\usepackage{xcolor}

%---------------- Header/Footer ----------------%
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\pagestyle{fancy}
\fancyhf{}
\rhead{NLP in Python — Spring (Modules 3–4)}
\lhead{Assoc.\ Prof.\ Soroosh Shalileh}
\rfoot{\thepage}

%---------------- Title ----------------%
\title{\textbf{Natural Language Processing (NLP) in Python}\\
\large Master’s in Data Science -- HSE University}
\author{Instructor: Assoc.\ Prof.\ Soroosh Shalileh \\
Email: \texttt{sshalileh@hse.ru}}
\date{Spring Semester (Modules 3–4)}

\begin{document}
\maketitle

\section*{Course Duration and Meeting Pattern}
\begin{itemize}[leftmargin=1.25em]
  \item \textbf{Length:} 13 teaching (educational) weeks across Spring Modules 3–4 (typ.\ January--June).
  \item \textbf{Contact hours:} 1 \emph{lecture} (two hours) $+$ 1 \emph{lab/seminar} (two hours).
  \item \textbf{Student workload:} 5--10 hours/week for readings, homework, and project work.
\end{itemize}

\section*{Course Description}
This course progresses from essential NLP foundations to state-of-the-art methods with an emphasis on \emph{advanced neural} and \emph{LLM-centric} techniques. We build core skills (tokenization, classical text classification, embeddings) and then dive into modern topics: neural networks for NLP, large language models (LLMs), transformers, masked LMs (BERT), retrieval-augmented generation (RAG), instruction tuning, machine translation, and recurrent models. Python labs use \texttt{PyTorch} and \texttt{HuggingFace Transformers}, with supporting use of \texttt{NLTK}/\texttt{spaCy}/\texttt{scikit-learn}.

\section*{Main Textbook}
\textbf{Daniel Jurafsky and James H.\ Martin}, \emph{Speech and Language Processing, 3rd ed.\ (online draft)}. Latest release: Aug 24, 2025.\\
\url{https://web.stanford.edu/~jurafsky/slp3/}

\section*{Course Outcomes}
By the end, students can:
\begin{enumerate}[leftmargin=1.25em]
  \item Implement classical and neural NLP models in Python.
  \item Fine-tune transformer models for classification and sequence labeling.
  \item Build RAG pipelines and evaluate IR components.
  \item Explain instruction tuning/align\-ment and masked vs.\ causal LMs.
\end{enumerate}

\section*{Weekly Schedule (Readings from Jurafsky \& Martin, SLP3)}
\subsection*{Module 3 — Foundations}
\begin{enumerate}[label=\textbf{Week \arabic*.}, leftmargin=*]
  \item \textbf{Intro to NLP, course Logistics - I}.
  Tasks, datasets, evaluation, ethics.
  Words \& Tokens, Edit Distance. Unicode, normalization, and tokenization. \\
        \emph{Reading:} Ch.\ 2 (\emph{Words and Tokens}).   
    \item \textbf{Intro to NLP - II}.
    Tasks, datasets, evaluation, ethics.
    Words \& Tokens, Edit Distance. Unicode, normalization, and tokenization. \\
    \emph{Reading:} Ch.\ 2 (\emph{Words and Tokens}). 
  \item \textbf{N-gram Language Models.} MLE, smoothing, perplexity.\\
        \emph{Reading:} Ch.\ 3 (\emph{N-gram Language Models}).
  \item \textbf{Logistic Regression for Text Classification.} Features for text, calibration, errors.\\
        \emph{Reading:} Ch.\ 4 (\emph{Logistic Regression and Text Classification}).
  \item \textbf{Embeddings (Static).} Distributional semantics, word2vec, GloVe; intrinsic eval.\\
        \emph{Reading:} Ch.\ 5 (\emph{Embeddings}).
  \item \textbf{Neural Networks (Feedforward) for NLP.} From features to learned reps.\\
        \emph{Reading:} Ch.\ 6 (\emph{Neural Networks}).
  \item \textbf{Large Language Models (Overview).} Causal LMs, sampling, scaling laws, safety.\\
        \emph{Reading:} Ch.\ 7 (\emph{Large Language Models}). \\
\end{enumerate}

\textbf{End of 3rd module:} mini-Project 1: End-to-end pipeline (tokenization $\rightarrow$ features/embeddings $\rightarrow$ classifier). In-class lightning talks.

\subsection*{Module 4 — Advanced Topics; emphasis on modern methods)}
\begin{enumerate}[label=\textbf{Week \arabic*.}, start=8, leftmargin=*]
    \item \textbf{Transformers (I):} Self-attention, encoder/decoder, positional encoding, scaling (it time permits, o.w will be shifted to next module).\\
        \emph{Reading:} Ch.\ 8 (\emph{Transformers}).
        
  \item \textbf{Transformers (II) \& Masked LMs (BERT).} Encoders, pretraining objectives, finetuning.\\
        \emph{Reading:} Ch.\ 10 (\emph{Masked Language Models}).
        
  \item \textbf{Retrieval \& RAG (I).} Dense/sparse retrieval, indexing, retrieval-augmented QA.\\
        \emph{Reading:} Ch.\ 11 (\emph{Information Retrieval and Retrieval-Augmented Generation}).

  \item \textbf{Instruction Tuning \& Alignment.} SFT, DPO, preference data, test-time compute.\\
        \emph{Reading:} Ch.\ 9 (\emph{Post-training: Instruction Tuning, Alignment, \& Test-Time Compute}).
  
  \item \textbf{Reserved.} Devoted to adjust the course plan and unexpected shifts. If we manage to cover all planned the materials, some optionl subject will be covered.   

  % \item \textbf{Machine Translation.} Encoder–decoder, attention, evaluation (BLEU/COMET), domain adaptation.\\
  %       \emph{Reading:} Ch.\ 12 (\emph{Machine Translation}).

  % \item \textbf{RNNs \& LSTMs (Legacy/Complementary).} Unrolling, BPTT, seq2seq; where RNNs remain useful.\\
  %       \emph{Reading:} Ch.\ 13 (\emph{RNNs and LSTMs}).

\item \textbf{Final Project Presentations.} Demo, report, and reflection.

        
\end{enumerate}


\subsection*{Optional Classes (if time permits)}
\begin{itemize}[leftmargin=1.25em]
  \item \textbf{Optional A:} Sequence Labeling and IE (Applications). NER/POS (BIO tagging), span extraction; IE basics.\\ 
  
  \emph{Reading:} Ch.\ 17 (\emph{Sequence Labeling for POS and NER}); Ch.\ 20 (\emph{Information Extraction}) \emph{(selected sections)}.
        
  \item \textbf{Optional B:} Interpretability for NLP (saliency, probing, LIME/SHAP, adversarial examples).
  \item \textbf{Optional C:} Advanced RAG systems (chunking strategies, reranking, evaluation pitfalls, caching).
\end{itemize}


\section*{Projects}
\textbf{Mini-Project 1:} your proposed subject (short paper + 5-min talk).\\
\textbf{Final Project:} your proposed subject approved by insructure. 
Deliverables: repo, 4–6 page report, 10-min demo.

\section*{Evaluation Policy}
\begin{tabular}{ll}
\textbf{Component} & \textbf{Weight} \\
\hline
Homework (7) & 35\% \\
Mini-Project 1 & 20\% \\
Final Project & 30\% \\
Participation (labs, discussions, peer feedback) & 10\% \\
Short quiz/reading checks (occasional) & 5\% \\
\end{tabular}



\section*{Submission Guidelines}

\subsection*{Homework Submissions}
\begin{itemize}[leftmargin=1.25em]
  \item All homework assignments must be submitted \textbf{together as a single \texttt{.zip} file} by \textbf{24th April}.
  \item The zip file must be named using the following convention: \texttt{Firstname\_Lastname\_NLP\_HW.zip}.
  \item Inside the zip, each homework should be in its own folder named \texttt{HW1}, \texttt{HW2}, \ldots,.
  \item Each homework folder must contain:
  \begin{itemize}
    \item A \textbf{Jupyter notebook} (\texttt{.ipynb}) or Python script(s) with clean, runnable code.
    \item A \textbf{short written report} (PDF, or within the notebook) explaining your approach, results, and analysis.
    \item A \texttt{README.txt} briefly describing how to run your code and any dependencies.
  \end{itemize}
  \item Submit the zip file via email to \texttt{sshalileh@hse.ru} with the subject line \texttt{[NLP HW] Firstname Lastname}.
  \item \textcolor{red}{\textbf{Reminder:} For each homework and each section, you must clearly state the problem. Submissions that fail to do so will receive \emph{zero} for that assignment. Submissions which do not adhere to the structure above will be penalized accordingly.}
\end{itemize}

\subsection*{Mini-Project Submission}
\begin{itemize}[leftmargin=1.25em]
  \item Submit a \textbf{GitHub repository}\footnote{The zip file of of the repository is also acceptable.} containing all code, data references, and a \texttt{README} describing how to reproduce your results.
  \item Include a \textbf{short paper} (2--3 pages, ACL two-column format) as a PDF in the repository root.
  \item Be prepared to give a \textbf{5-minute in-class presentation} (slides optional but recommended).
  \item Share the repository link by emailing \texttt{sshalileh@hse.ru} with the subject line \texttt{[NLP Mini-Project] Firstname Lastname} by 26 March 2026.
\end{itemize}

\subsection*{Final Project Submission}
\begin{itemize}[leftmargin=1.25em]
  \item Submit a \textbf{GitHub repository} containing all code, data references, and a \texttt{README} describing how to reproduce your results.
  \item Include a \textbf{project report} (4--6 pages, ACL two-column format) as a PDF in the repository root.
  \item Be prepared to give a \textbf{10-minute in-class demo/presentation}.
  \item Share the repository link by emailing \texttt{sshalileh@hse.ru} with the subject line \texttt{[NLP Final Project] Firstname Lastname} by \textbf{10 May}.
  \item \textcolor{red}{\textbf{Reminder:} editing/pushing repository after the deadline will be considered as missing the deadline and you will loss the entire score.
\end{itemize}

\section*{Homework Assignments (5 total) - Deadline: 24 April 2026, in one collection i.e. \textbf{one zip file}}

\textcolor{red}{\textbf{Attention: \newline 
\noindent For each homework and each section of it, you should clearly state the problem. 
Those who fail to clearly and cleanly explain the problem will lose the ENTIRE score.}}


\begin{enumerate}[leftmargin=1.25em]
  \item \textbf{HW1 (Foundations - Ch.\ 2--3):} 
  \\ Normalization \& tokenization: 
  select a data set of your choice, conduct a brief exploratory data analysis, decide an arbitrary task you are supposed to implement, and apply the text normalization and tokenization as discussed in the classed. 
  \\ Implement either Edit Distance or Byte-pairing algorithm and apply it to you data set and analysis the output and comment on it. 
  \item \textbf{HW2 (Classical Text classification - Ch.\ 4):} Construct at least two variant of N-gram (justify your choice) and apply logistic regression on them and evaluate the results using proper metrics, justify your choice and analyze them.
  \item \textbf{HW3 (Embeddings - Ch.\ 5):} Train word2vec/GloVe on a small corpus; visualize neighborhoods; intrinsic vs.\ extrinsic evaluation. .
  \item \textbf{HW4 (Neural Basics - Ch.\ 6):} \\ Option 1: Feedforward neural network text classifier with learned embeddings.
  \\ Option 2: Use a feedforward neural network to obtained an embedding. 
  \\ For each option the choice of each component should be clarified clearly and justified. 
  % \item \textbf{HW5 (MLM Finetuning):} 
  % Finetune a masked-LM encoder (e.g., \texttt{bert-base-uncased}) for NER or sentiment; ablations (Ch.\ 9).


  %%
  \item \textbf{HW5 (Transformers \& Causal Language Modeling):} \newline
  \textbf{Objective:}
    This assignment focuses on understanding self-attention, transformer architecture, and causal language modeling. Students will implement core components, finetune a pretrained model, and analyze attention behavior.
    
    \vspace{0.5em}
    
    \textbf{Part 1 — Conceptual Questions (20 pts)}
    
    \begin{enumerate}
    \item Explain self-attention. What are queries, keys, and values? Why is attention a weighted sum?
    
    \item What is causal masking and why is it necessary in language models?
    
    \item Why are multiple attention heads used instead of a single head?
    
    \item What is the role of residual connections in transformer models?
    
    \vspace{0.5em}

    \end{enumerate}
    
    \textbf{Part 2 — Implement Self-Attention (25 pts)}
    
    Implement a single-head self-attention layer from scratch (no deep learning libraries for attention).
    
    Given input matrix $X \in \mathbb{R}^{n \times d}$:
    
    \begin{itemize}
    \item Compute $Q = XW_Q$, $K = XW_K$, $V = XW_V$
    \item Compute attention scores:
    \[
    A = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right)
    \]
    \item Apply causal masking
    \item Output:
    \[
    \text{Attention}(X) = AV
    \]
    \end{itemize}
    
    \textbf{Deliverables:}
    \begin{itemize}
    \item Code implementation
    \item Test on a toy example (e.g., 3 tokens)
    \end{itemize}
    
    \vspace{0.5em}
    
    \textbf{Part 3 — Causal Language Model Finetuning (30 pts)}
    
    Finetune a pretrained causal language model (e.g., \texttt{gpt2}) using Hugging Face Transformers. Quantized and small models like Gemma is also acceptable. 
    
    \textbf{Dataset:}
    Choose a small domain-specific dataset (e.g., tweets, reviews, news headlines).
    
    \textbf{Tasks:}
    \begin{itemize}
    \item Finetune the model
    \item Generate text from prompts
    \item Report training loss and example outputs
    \end{itemize}
    
    \vspace{0.5em}
    
    \textbf{Part 4 — Attention Analysis (15 pts)}
    
    Analyze attention patterns in the finetuned model.
    
    \begin{itemize}
    \item Visualize attention weights for at least 2 layers
    \item Select example sentences and inspect which tokens attend to which
    \item Discuss whether different heads capture different patterns
    \end{itemize}
    
    \vspace{0.5em}
    
    \textbf{Part 5 — Ablation Study (10 pts)}
    
    Run at least two of set of experiments, for instance, Reduce number of attention heads, change context window size, learning rate, number of epochs etc., and 
        
    Discuss their impact on model's performance. 
    
    \vspace{0.5em}
    
    % \textbf{Part 5 — Generation Behavior (10 pts)}
    
    % Compare text generation under different decoding strategies:
    
    % \begin{itemize}
    % \item greedy decoding
    % \item sampling with temperature
    % \item top-k or top-p sampling
    % \end{itemize}
    
    % Provide examples and brief analysis.
    
    % \vspace{0.5em}
    
    \textbf{Submission Requirements:}
    \begin{itemize}
    \item Code (Jupyter notebook or Python scripts)
    \item Short report (3–5 pages) explaining results
    \item Plots or visualizations where applicable
    \end{itemize}
    
    \vspace{0.5em}
    
    \textbf{Optional Bonus (+10 pts):}
    \begin{itemize}
    \item Implement multi-head attention from scratch
    \item Compare with single-head version
    \end{itemize}
  %%
  
  % \item \textbf{HW6 (RAG Mini-System):} TBD
  % Build a retrieval-augmented QA pipeline (BM25 or dense retriever $+$ reader); report latency/accuracy tradeoffs (Ch.\ 11).

\end{enumerate}




\section*{Software and Tools}
Python 3.x; Jupyter/Colab; \texttt{NLTK}, \texttt{spaCy}, \texttt{scikit-learn}, \texttt{gensim}; \texttt{PyTorch}; \texttt{HuggingFace Transformers}; optional \texttt{Weights\&Biases}/TensorBoard. GitHub Classroom for submissions is recommended.



\end{document}
